A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-23 Manuscript/PNAS-Adams-03.tex
vsn 3

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Evolutionary biologists wish to quantify and compare the strength of
phylogenetic signal across datasets, but analytical tools for these
comparisons are generally lacking. Here we develop a standardized effect
size, \(Z_\kappa\), which measures the strength of phylogenetic signal
on a common statistical scale. We also provide a test statistic,
\(\hat{Z}_{12}\), for comparing the strength of phylogenetic signal
across datasets. We find that two commonly used parameters (Pagel's
\(\lambda\) and Blomberg's \(\kappa\)), not converted to effect sizes,
are unsuitable for this purpose. Our effect-size procedure enables
biologists to quantitatively address hypotheses that compare the
strength of phylogenetic signal between various phenotypic traits, even
when those traits are found in different evolutionary lineages or have
different units or scales.}

@@

\begin{abstract}
Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, however, analytical tools for comparing the
strength of that signal across traits remain largely underdeveloped.
Here we evaluate the efficacy of Pagel's \(\lambda\) to correctly
estimate the strength of phylogenetic signal in phenotypic traits across
a range of input values. We find that \(\lambda\) behaves as a Bernoulli
random variable, where estimates are increasingly skewed at larger and
smaller input levels of phylogenetic signal. Further, the precision of
\(\lambda\) varies with input signal. Another measure, Blomberg's
\(\kappa\), is more consistent across a range of tree sizes, and
exhibits a positive relationship with input levels of phylogenetic
signal. However, that relationship is decidedly nonlinear. Thus, neither
\(\lambda\) nor \(\kappa\) are suitable as effect sizes for measuring
the strength of phylogenetic signal, and comparing that signal across
datasets. As an alternative, we propose a standardized effect size based
on \(\kappa\), (\(Z_\kappa\)), which measures the strength of
phylogenetic signal more reliably than does \(\lambda\), and places that
signal on a common scale for statistical comparison. We develop tests
based on \(Z_\kappa\) to provide a mechanism for formally comparing the
strength of phylogenetic signal across datasets, in much the same manner
as effect sizes may be used to summarize patterns in quantitative
meta-analysis. Our approach extends the phylogenetic comparative toolkit
to address hypotheses that compare the strength of phylogenetic signal
between various phenotypic traits, even when those traits are found in
different evolutionary lineages or have different units or scales.
\end{abstract}

@@

Another widely used measure of phylogenetic signal is Blomberg's
\(\kappa\) (12), which characterizes phylogenetic signal as the ratio of
observed trait variation to the amount of variation expected under
Brownian motion. Blomberg's \(\kappa\) can be treated as a test
statistic by employing a permutation test to generate its sampling
distribution (12, 16) for determining whether significant phylogenetic
signal is present in data. Both \(\lambda\) and \(\kappa\) seem
intuitive to interpret, as a value of \(0\) for both corresponds to no
phylogenetic signal, while a value of \(1\) corresponds to the amount of
phylogenetic signal expected under Brownian motion. Thus, it is tempting
to regard both \(\lambda\) and \(\kappa\) as descriptive statistics that
measure the relative strength of phylogenetic signal, providing an
estimate of its magnitude for comparison.

The appeal of Pagel's \(\lambda\) and Blomberg's \(\kappa\) as
descriptive statistics is that they provide a basis for interpreting
``weak'' versus ``strong'' phylogenetic signal; i.e., small versus large
values of \(\hat{\lambda}\) or \(\kappa\), respectively, in a
comparative sense (28--30). Nonetheless, an important question that has
yet to be considered is whether such comparisons are analytically
appropriate, and whether these statistics are, or can be, converted to
effect sizes for comparative analyses across datasets. To be statistics
representing phylogenetic signal, they should have reliable
distributional properties, which could be revealed with simulation
experiments. For instance, as a proportional random variable bounded by
\(0\) and \(1\), we might expect that \(\hat{\lambda}\) is a random
variable that follows the distribution of a Bernoulli probability
parameter (31); i.e., branch lengths in a tree are scaled proportionally
to the probability that data arise from a BM process. Given a known
\(\lambda\) value used to generate random data on a tree, we would also
expect that the mean of an empirical sampling distribution of
\(\hat{\lambda}\) would approximately equal \(\lambda\); the dispersion
of \(\hat{\lambda}\) would be largest at intermediate values of
\(\lambda\), \(\hat{\lambda}\) would be predictable over the range of
\(\lambda\) with respect to tree size; the distribution of
\(\hat{\lambda}\) would be symmetric at intermediate values of
\(\lambda\) and more skewed toward values of 0 or 1; and that the
distribution of \(\hat{\lambda}\) will be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1 (31). Prior work (18) seems to support some of these conjectures,
based superficially on statistical moments for a given tree size (mean,
variance, skewness, and kurtosis; see Fig. 2 of ref. (18)). However,
because the ``strength of Brownian motion'' was simulated as a varied
weighted-average of data simulated on trees with \(\lambda=0\) and
\(\lambda=1\) and not as prescribed values of \(\lambda\) (18),
interpretation of these patterns is challenging.

By contrast, for Blomberg's \(\kappa\), which is positively unbounded,
we might expect that for any \(\lambda\) used to generate data,
estimates of \(\kappa\) might be a random variable that follows a normal
distribution, with values distributed symmetrically about the input
value (31). This attribute seemed less reasonable based on the
simulations performed by Münkemüller et al.~(18), which suggested that
distributions were positively skewed and that Blomberg's \(\kappa\)
might not behave as a statistic that follows a normal distribution.
However, because their simulations used a weighted combination of
simulated phylogenetic signal strengths, strong inferences are not
possible (and distributional attributes were not the intended result of
their simulations). Thus, for both Pagel's \(\lambda\) or Blomberg's
\(\kappa\), evaluation of statistical moments across a range of
\(\lambda\) used to generate data would be valuable for adjudicating the
reliability of these statistics as effect sizes. Furthermore, the
expected values of these statistics appear to vary with tree size (18),
making comparisons across studies challenging. Therefore, transformation
of these statistics into \(Z\)-scores would allow evaluation of the
efficacy of each statistic to yield effect sizes that could be used for
comparisons of the strength of phylogenetic signal across traits and
lineages.

Here we use simulation experiments to compare the distributional
attributes of \(\hat{\lambda}\) and \(\kappa\), plus their effect sizes
(\(Z\)-scores), across a range of tree size and phylogenetic signal
strength. We find that estimates of \(\hat{\lambda}\) are increasingly
skewed at larger and smaller input levels of phylogenetic signal and at
smaller tree sizes, vary widely for a given input value of \(\lambda\),
and that the precision of \(\hat{\lambda}\) is not constant across its
range. By contrast, estimates of \(\kappa\) are more consistent across
tree sizes, and are normally distributed across the range of input
levels of \(\lambda\), making \(\kappa\) a more reliable statistic. We
then propose an effect size based on \(\kappa\), (\(Z_{\kappa}\)), which
provides consistent estimates of the strength of phylogenetic signal
across tree sizes and signal strength, and facilitates quantitative
comparisons of the relative strength of phylogenetic signal across
datasets.

@@

\textbf{Lambda (\(\lambda\)) estimates of phylogenetic signal are
innacurate.} Computer simulations reveal that for \(\hat{\lambda}\), the
distributional expectations of a Bernoulli variable were mostly upheld.
First, the mean value of \(\hat{\lambda}\) increases as \(\lambda\)
increases, but it is negatively-biased (particularly for small tree
sizes), and is consistently less than the input \(\lambda\) value across
most of its range (Fig. 1 black line). Second, the standard deviation of
\(\hat{\lambda}\) is largest at intermediate values of \(\lambda\) and
smallest at extreme values (Fig.1 red line), implying that the precision
in estimating \(\lambda\) varies across the range of input values.
Additionally, standard deviations of \(\hat{\lambda}\) are negatively
associated with tree size, and for tress of 128 species or less,
\(\hat{\lambda}\) are quite variable, except for cases when \(\lambda\)
is near or equal to \(1\). Third, the distributions of \(\hat{\lambda}\)
are not normal across its range, but become increasingly skewed at more
extreme values of \(\lambda\) (Fig. 1 blue line). For small tree sizes,
it is also clear that distributions are more platykurtic at intermediate
values of \(\hat{\lambda}\). Taken together these results reveal that
\(\hat\lambda\) inconsistently estimates phylogenetic signal, both
across tree sizes and across the range of input values. Additional
simulations (Supplemental Information) reveal that incorporating
\(\hat\lambda\) in PGLS ANOVA and regression does not adversely affect
the statistical properties of PGLS parameter estimation or model
evaluation (type I error, power, bias in coefficients). Thus, it is
reasonable to incorporate \(\hat\lambda\) in PGLS as a parameter for
tuning the degree of phylogenetic signal in the dependent variables
during the analysis. However, the statistical properties shown in Fig. 1
demonstrate that \(\lambda\) is unsuitable as an effect size for
measuring the strength of phylogenetic signal in data, and thus
\(\lambda\) should not be used for comparing phylogenetic signal across
datasets.

@@

\textbf{Kappa (\(\kappa\)) estimates of phylogenetic signal are more
reliable.} Simulation results for \(\hat\kappa\) demonstrate that
\(\hat\kappa\) displays better statistical properties. First, as
expected, mean values of \(\hat\kappa\) increase with increasing signal
(\(\lambda\)) irrespective of tree size, though the increase does not
scale linearly with input levels of phylogenetic signal (Fig. 2 black
line). Additionally, the standard deviation of \(\hat\kappa\) is
consistent across tree sizes (Fig. 2 red line), and while it increases
with \(\lambda\), it is always less than the corresponding mean. This
finding is perhaps unsurprising, as \(\hat\kappa\) is lower-bounded by
0, and is never large for small values of \(\lambda\). Importantly,
\(\hat\kappa\) is normally distributed across the range of input
\(\lambda\), and remains consistent in this pattern regardless of tree
size (Fig. 2 blue line). This result differs from those of (18), where
the skewing appears to be the result of combining random values
generated independently, rather than being a property of \(\kappa\)
itself. Overall, these findings reveal that while \(\kappa\) is more
reliable as an estimate of phylogenetic signal, the non-linear scaling
with input signal implies that it should not be considered an effect
size that measures the strength of phylogenetic signal on a common scale
for comparison across datasets.

\textbf{Effect sizes from \(\kappa\) (\(Z_{\kappa}\)) better
characterize phylogenetic signal.} To measure the strength of
phylogenetic signal on a common scale, we propose effect sizes
(Z-scores) for both \(\lambda\) and \(\kappa\). Statistically, a
standardized effect size may be found as:

@@

Formalizing the suggestion of Adams and Collyer (39), an effect size for
\(\kappa\) may be found as:

\begin{align}
    Z_\kappa=\frac{\kappa_{obs}-\hat\mu_{\kappa}}{\hat\sigma_{\kappa}},
\end{align}

where \(\kappa_{obs}\) is the observed phylogenetic signal, and
\(\hat\mu_\kappa\) and \(\hat\sigma_\kappa\) are the mean and standard
deviation of the empirical sampling distribution of \(\kappa\) obtained
via permutation. The empirical sampling distribution of \(\kappa\) is
first transformed via a Box-Cox transformation to better adhere to the
assumption of normality.

For \(\lambda\), deriving an effect size is more challenging, as
\(\lambda\) does not have a sampling distribution from which the
standard error and confidence intervals may be obtained, and estimates
from the Hessian matrix from PGLS are unreliable (23). Confidence
intervals are therefore generated for the values of \(\lambda\) that
intersect the log-likehihood profile for corresponding percentiles of
the \(\chi^2\) distribution used to compare the putative model to a null
model with \(\lambda = 0\) {[}\textbf{add ref: MLC thinks Boettiger
paper?}{]}. Thus, an effect size for \(\lambda\) may be found as:

@@

Here we evaluate the ability of \(Z_{\lambda}\) and \(Z_{\kappa}\) to
characterize known levels of phylogenetic signal. Both \(Z_{\lambda}\)
and \(Z_{\kappa}\) are associated with input phylogenetic signal
(\(\lambda\)), indicating that both statistics capture the observed
signal (Fig. 3). However, effect sizes from \(\hat{\lambda}\) made
little sense, as they are more strongly associated with tree size than
they are with the actual phylogenetic signal in the data (Fig. 3). By
contrast, \(Z_{\kappa}\) is much more consistent across tree sizes, and
increases more linearly with increasing levels of phylogenetic signal.
Additionally, \(Z_{\kappa}\) exhibits a much stronger association with
phylogenetic signal strength as compared to tree size (Fig. 3), and its
standard deviation across input signal is more consistent. This implies
that similar levels of precision are found with \(Z_{\kappa}\) across
the range of input values. Thus between the two statistics,
\(Z_{\kappa}\) is a more reliable measure of the strength of
phylogenetic signal, and may be used to compare levels of phylogenetic
signal across datasets.

\begin{figure}
\centering
\includegraphics[width=8.25cm]{fig.3.png}
\caption{Response of effect sizes \(Z_{\lambda}\) and \(Z_{\kappa}\) to increasing strength of Brownian motion.{}}
\end{figure}

@@

\begin{align}
  \hat{Z}_{12}=\frac{\lvert{(\kappa_{1}-\hat\mu_{\kappa_1})-(\kappa_{2}-\hat\mu_{\kappa_2})}\rvert}{\sqrt{\hat\sigma^2_{\kappa_1}+\hat\sigma^2_{\kappa_2}}} = \frac{\lvert Z_{\kappa_1} - Z_{\kappa_2}\rvert}{\sqrt{2}}
\end{align}

where \(\kappa_1\), \(\kappa_2\), \(\hat\mu_{\kappa_1}\),
\(\hat\mu_{\kappa_2}\), \(\hat\sigma_{\kappa_1}\), and
\(\hat\sigma_{\kappa_2}\) are as defined above for equation 2. The right
side of the equation illustrates that if \(Z_\kappa\) has already been
calculated for two sampling distributions as in equation 2, the sampling
distributions have unit variance for each of the \(Z_\kappa\)
statistics. Estimates of significance of \(\hat{Z}_{12}\) may be
obtained from a standard normal distribution. Typically,
\(\hat{Z}_{12}\) is considered a two-tailed test, however directional
(one-tailed) tests may be specified should the empirical situation
require it (36, 38).

To demonstrate the utility of \(\hat{Z}_{12}\), we compared
\(Z_{\kappa}\) for two ecologically-relevant traits in plethodontid
salamander (Fig. 4): surface area to volume ratios (SA:V) and relative
body width (\(\frac{BW}{SVL}\)) (40, 41). While both traits contained
significant phylogenetic signal, tests based on \(\hat{Z}_{12}\)
revealed that the degree of phylogenetic signal was significantly
stronger in SA:V (\(\hat{Z}_{12}=4.13\); \(P=0.000036\): Fig. 4).
Biologically, this observation may be interpreted by the fact that
tropical species -- which form a monophyletic group within plethodontids
-- display greater variation in SA:V, which covaries with disparity in
their climatic niches (41). Thus, greater phylogenetic signal in SA:V is
to be expected.

\begin{figure}
\centering
\includegraphics{fig.4.png}
\caption{(A) Linear measures for relative body size, and regions of the
body used to estimate surface area to volume (SA:V) ratios. (B)
Permutation distributions of phylogenetic signal for SA:V and
\(\frac{BW}{SVL}\), with observed values shown as vertical bars. (C)
Effect sizes (\(Z_\kappa\)) for SA:V and \(\frac{BW}{SVL}\), with their
95\% confidence intervals (CI not standardized by \(\sqrt(n)\)).{}}
\end{figure}

@@

As an alternative, we described a standardized effect size (\(Z\)) for
assessing the strength of phylogenetic signal. \(Z\) expresses the
magnitude of phylogenetic signal as a standard normal deviate, which is
easily interpretable as the strength of phylogenetic signal relative to
the mean. We applied this concept to both \(\lambda\) and \(\kappa\),
and found that \(Z_\kappa\) was a better estimate of the strength of
phylogenetic signal in phenotypic data. First, values of \(Z_\kappa\)
more accurately tracked known changes in the magnitude of phylogenetic
signal, as demonstrated by the linear relationship between \(Z_\kappa\)
and input signal. Additionally, the precision of \(Z_{\kappa}\) was more
consistent across the range of input levels of phylogenetic signal.
Thus, \(Z_\kappa\) is a more reliable measure of the relative strength
of phylogenetic signal, and places that effect on a common and
comparable scale. We therefore recommend that future studies interested
in evaluating the strength of phylogenetic signal incorporate
\(Z_\kappa\) as a statistical measure of this effect.

Next we proposed a two-sample test (\(\hat{Z}_{12}\)), which provides a
formal statistical procedure for determining whether the strength of
phylogenetic signal is greater in one phenotypic trait as compared to
another. Prior studies have summarized patterns of variation in
phylogenetic signal across datasets using summary test values, such as
\(\kappa\) (12). However, because \(\kappa\) does not scale linearly
with input levels of phylogenetic signal (Fig. 2), and its variance
increases with increasing strength of phylogenetic signal (18, 20), it
should not be considered an effect size that measures the strength of
phylogenetic signal on a common scale. By contrast, standardizing
\(\kappa\) to \(Z_\kappa\) via equation 2 alleviates these concerns, and
facilitates formal statistical comparisons of the strength of signal
across datasets. Thus when viewed from this perspective, the approach
developed here aligns well with other statistical approaches such as
meta-analysis (32, 42, 43), where summary statistics across datasets are
converted to standardized effect sizes for subsequent ``higher order''
statistical summaries or comparisons. As such, our approach enables
evolutionary biologists to quantitatively examine the relative strength
of phylogenetic signal across a wide range of phenotypic traits, and
thus opens the door for future discoveries that inform on how phenotypic
diversity accumulates in macroevolutionary time across the tree of life.

>>>>

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Evolutionary biologists wish to quantify and compare the strength of
phylogenetic signal across datasets, but analytical tools for these
comparisons are generally lacking. Here we develop a standardized effect
size, \(Z_K\), which measures the strength of phylogenetic signal on a
common statistical scale. We also provide a test statistic,
\(\hat{Z}_{12}\), for comparing the strength of phylogenetic signal
across datasets. We find that two commonly used parameters (Pagel's
\(\lambda\) and Blomberg's \emph{K}), not converted to effect sizes, are
unsuitable for this purpose. Our effect-size procedure enables
biologists to quantitatively address hypotheses that compare the
strength of phylogenetic signal between various phenotypic traits, even
when those traits are found in different evolutionary lineages or have
different units or scales.}

@@

\begin{abstract}
Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, however, analytical tools for comparing the
strength of that signal across traits remain largely underdeveloped.
Here we evaluate the efficacy of Pagel's \(\lambda\) to correctly
estimate the strength of phylogenetic signal in phenotypic traits across
a range of input values. We find that \(\lambda\) behaves as a Bernoulli
random variable, where estimates are increasingly skewed at larger and
smaller input levels of phylogenetic signal. Further, the precision of
\(\lambda\) varies with input signal. Another measure, Blomberg's
\emph{K}, is more consistent across a range of tree sizes, and exhibits
a positive relationship with input levels of phylogenetic signal.
However, that relationship is decidedly nonlinear. Thus, neither
\(\lambda\) nor \emph{K} are suitable as effect sizes for measuring the
strength of phylogenetic signal, and comparing that signal across
datasets. As an alternative, we propose a standardized effect size based
on \emph{K}, (\(Z_K\)), which measures the strength of phylogenetic
signal more reliably than does \(\lambda\), and places that signal on a
common scale for statistical comparison. We develop tests based on
\(Z_K\) to provide a mechanism for formally comparing the strength of
phylogenetic signal across datasets, in much the same manner as effect
sizes may be used to summarize patterns in quantitative meta-analysis.
Our approach extends the phylogenetic comparative toolkit to address
hypotheses that compare the strength of phylogenetic signal between
various phenotypic traits, even when those traits are found in different
evolutionary lineages or have different units or scales.
\end{abstract}

@@

Another widely used measure of phylogenetic signal is Blomberg's
\emph{K} (12), which characterizes phylogenetic signal as the ratio of
observed trait variation to the amount of variation expected under
Brownian motion. Blomberg's \emph{K} can be treated as a test statistic
by employing a permutation test to generate its sampling distribution
(12, 16) for determining whether significant phylogenetic signal is
present in data. Both \(\lambda\) and \emph{K} seem intuitive to
interpret, as a value of \(0\) for both corresponds to no phylogenetic
signal, while a value of \(1\) corresponds to the amount of phylogenetic
signal expected under Brownian motion. Thus, it is tempting to regard
both \(\lambda\) and \emph{K} as descriptive statistics that measure the
relative strength of phylogenetic signal, providing an estimate of its
magnitude for comparison.

The appeal of Pagel's \(\lambda\) and Blomberg's \emph{K} as descriptive
statistics is that they provide a basis for interpreting ``weak'' versus
``strong'' phylogenetic signal; i.e., small versus large values of
\(\hat{\lambda}\) or \emph{K}, respectively, in a comparative sense
(28--30). Nonetheless, an important question that has yet to be
considered is whether such comparisons are analytically appropriate, and
whether these statistics are, or can be, converted to effect sizes for
comparative analyses across datasets. To be statistics representing
phylogenetic signal, they should have reliable distributional
properties, which could be revealed with simulation experiments. For
instance, as a proportional random variable bounded by \(0\) and \(1\),
we might expect that \(\hat{\lambda}\) is a random variable that follows
the distribution of a Bernoulli probability parameter (31); i.e., branch
lengths in a tree are scaled proportionally to the probability that data
arise from a BM process. Given a known \(\lambda\) value used to
generate random data on a tree, we would also expect that the mean of an
empirical sampling distribution of \(\hat{\lambda}\) would approximately
equal \(\lambda\); the dispersion of \(\hat{\lambda}\) would be largest
at intermediate values of \(\lambda\), \(\hat{\lambda}\) would be
predictable over the range of \(\lambda\) with respect to tree size; the
distribution of \(\hat{\lambda}\) would be symmetric at intermediate
values of \(\lambda\) and more skewed toward values of 0 or 1; and that
the distribution of \(\hat{\lambda}\) will be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1 (31). Prior work (18) seems to support some of these conjectures,
based superficially on statistical moments for a given tree size (mean,
variance, skewness, and kurtosis; see Fig. 2 of ref. (18)). However,
because the ``strength of Brownian motion'' was simulated as a varied
weighted-average of data simulated on trees with \(\lambda=0\) and
\(\lambda=1\) and not as prescribed values of \(\lambda\) (18),
interpretation of these patterns is challenging.

By contrast, for Blomberg's \emph{K}, which is positively unbounded, we
might expect that for any \(\lambda\) used to generate data, estimates
of \emph{K} might be a random variable that follows a normal
distribution, with values distributed symmetrically about the input
value (31). This attribute seemed less reasonable based on the
simulations performed by Münkemüller et al.~(18), which suggested that
distributions were positively skewed and that Blomberg's \emph{K} might
not behave as a statistic that follows a normal distribution. However,
because their simulations used a weighted combination of simulated
phylogenetic signal strengths, strong inferences are not possible (and
distributional attributes were not the intended result of their
simulations). Thus, for both Pagel's \(\lambda\) or Blomberg's \emph{K},
evaluation of statistical moments across a range of \(\lambda\) used to
generate data would be valuable for adjudicating the reliability of
these statistics as effect sizes. Furthermore, the expected values of
these statistics appear to vary with tree size (18), making comparisons
across studies challenging. Therefore, transformation of these
statistics into \(Z\)-scores would allow evaluation of the efficacy of
each statistic to yield effect sizes that could be used for comparisons
of the strength of phylogenetic signal across traits and lineages.

Here we use simulation experiments to compare the distributional
attributes of \(\hat{\lambda}\) and \emph{K}, plus their effect sizes
(\(Z\)-scores), across a range of tree size and phylogenetic signal
strength. We find that estimates of \(\hat{\lambda}\) are increasingly
skewed at larger and smaller input levels of phylogenetic signal and at
smaller tree sizes, vary widely for a given input value of \(\lambda\),
and that the precision of \(\hat{\lambda}\) is not constant across its
range. By contrast, estimates of \emph{K} are more consistent across
tree sizes, and are normally distributed across the range of input
levels of \(\lambda\), making \emph{K} a more reliable statistic. We
then propose an effect size based on \emph{K}, (\(Z_K\)), which provides
consistent estimates of the strength of phylogenetic signal across tree
sizes and signal strength, and facilitates quantitative comparisons of
the relative strength of phylogenetic signal across datasets.

@@

\textbf{Lambda (\(\lambda\)) estimates of phylogenetic signal are
inaccurate.} Computer simulations reveal that for \(\hat{\lambda}\), the
distributional expectations of a Bernoulli variable were mostly upheld.
First, the mean value of \(\hat{\lambda}\) increases as \(\lambda\)
increases, but it is negatively-biased (particularly for small tree
sizes), and is consistently less than the input \(\lambda\) value across
most of its range (Fig. 1 black line). Second, the standard deviation of
\(\hat{\lambda}\) is largest at intermediate values of \(\lambda\) and
smallest at extreme values (Fig.1 red line), implying that the precision
in estimating \(\lambda\) varies across the range of input values.
Additionally, standard deviations of \(\hat{\lambda}\) are negatively
associated with tree size, and for tress of 128 species or less,
\(\hat{\lambda}\) are quite variable, except for cases when \(\lambda\)
is near or equal to \(1\). Third, the distributions of \(\hat{\lambda}\)
are not normal across its range, but become increasingly skewed at more
extreme values of \(\lambda\) (Fig. 1 blue line). For small tree sizes,
it is also clear that distributions are more platykurtic at intermediate
values of \(\hat{\lambda}\). Taken together these results reveal that
\(\hat\lambda\) inconsistently estimates phylogenetic signal, both
across tree sizes and across the range of input values. Additional
simulations (Supporting Information) reveal that incorporating
\(\hat\lambda\) in PGLS ANOVA and regression does not adversely affect
the statistical properties of PGLS parameter estimation or model
evaluation (type I error, power, bias in coefficients). Thus, it is
reasonable to incorporate \(\hat\lambda\) in PGLS as a parameter for
tuning the degree of phylogenetic signal in the dependent variables
during the analysis. However, the statistical properties shown in Fig. 1
demonstrate that \(\lambda\) is unsuitable as an effect size for
measuring the strength of phylogenetic signal in data, and thus
\(\lambda\) should not be used for comparing phylogenetic signal across
datasets.

@@

\textbf{Kappa (\emph{K}) estimates of phylogenetic signal are more
reliable.} Simulation results demonstrate that \emph{K} displays better
statistical properties. First, as expected, mean values of \emph{K}
increase with increasing signal (\(\lambda\)) irrespective of tree size,
albeit non-linearly (Fig. 2 black line). Additionally, the standard
deviation of \emph{K} is consistent across tree sizes (Fig. 2 red line),
and while it increases with \(\lambda\), it is always less than the
mean. This finding is perhaps unsurprising, as \emph{K} is lower-bounded
by 0, and is never large for small values of \(\lambda\). Importantly,
\emph{K} is normally distributed across the range of input \(\lambda\);
a consistent pattern regardless of tree size (Fig. 2 blue line). This
result differs from those of (18), where the skewing appears to be the
result of combining random values generated independently, rather than
being a property of \emph{K} itself. Overall, these findings reveal that
while \emph{K} is more reliable as an estimate of phylogenetic signal,
the non-linear scaling with input signal implies that it should not be
considered an effect size that measures the strength of phylogenetic
signal on a common scale for comparison across datasets.

\textbf{Effect sizes from \emph{K} (\(Z_K\)) better characterize
phylogenetic signal.} To measure the strength of phylogenetic signal on
a common scale, we propose effect sizes (Z-scores) for both \(\lambda\)
and \emph{K}. Statistically, a standardized effect size may be found as:

@@

Formalizing the suggestion of Adams and Collyer (39), an effect size for
\emph{K} may be found as:

\begin{align}
    Z_K=\frac{K_{obs}-\hat\mu_{K}}{\hat\sigma_{K}},
\end{align}

where \(K_{obs}\) is the observed phylogenetic signal, and \(\hat\mu_K\)
and \(\hat\sigma_K\) are the mean and standard deviation of the
empirical sampling distribution of \emph{K} obtained via permutation.
The empirical sampling distribution of \emph{K} is first transformed via
a Box-Cox transformation to better adhere to the assumption of
normality.

For \(\lambda\), deriving an effect size is more challenging, as
\(\lambda\) does not have a sampling distribution from which the
standard error and confidence intervals may be obtained, and estimates
from the Hessian matrix from PGLS are unreliable (23). Confidence
intervals are therefore generated for the values of \(\lambda\) that
intersect the log-likehihood profile for corresponding percentiles of
the \(\chi^2\) distribution used to compare the putative model to a null
model with \(\lambda = 0\) (40). Thus, an effect size for \(\lambda\)
may be found as:

@@

Simulations reveal that both \(Z_{\lambda}\) and \(Z_K\) are associated
with input phylogenetic signal (\(\lambda\)), indicating that both
statistics capture the observed signal (Fig. 3). However, effect sizes
from \(\hat{\lambda}\) made little sense, as they are more strongly
associated with tree size than they are with the actual phylogenetic
signal in the data (Fig. 3). By contrast, \(Z_K\) is much more
consistent across tree sizes, and increases more linearly with
increasing levels of phylogenetic signal. Additionally, \(Z_K\) exhibits
a much stronger association with phylogenetic signal strength as
compared to tree size (Fig. 3), and its standard deviation is more
consistent, implying similar levels of precision across the range of
input signal (Supporting Information). Thus, \(Z_K\) is a more reliable
measure of the strength of phylogenetic signal, and may be used to
compare levels of phylogenetic signal across datasets.

\begin{figure}
\centering
\includegraphics[width=8.25cm]{fig.3.png}
\caption{Response of effect sizes \(Z_{\lambda}\) and \(Z_K\) to increasing strength of Brownian motion.{}}
\end{figure}

@@

\begin{align}
  \hat{Z}_{12}=\frac{\lvert{(K_{1}-\hat\mu_{K_1})-(K_{2}-\hat\mu_{K_2})}\rvert}{\sqrt{\hat\sigma^2_{K_1}+\hat\sigma^2_{K_2}}} = \frac{\lvert Z_{K_1} - Z_{K_2}\rvert}{\sqrt{2}}
\end{align}

where \(K_1\), \(K_2\), \(\hat\mu_{K_1}\), \(\hat\mu_{K_2}\),
\(\hat\sigma_{K_1}\), and \(\hat\sigma_{K_2}\) are as defined above for
equation 2. The right side of the equation illustrates that if \(Z_K\)
has already been calculated for two sampling distributions, both have
unit variance for each of the \(Z_K\) statistics. Estimates of
significance of \(\hat{Z}_{12}\) may be obtained from a standard normal
distribution. Typically, \(\hat{Z}_{12}\) is considered a two-tailed
test, however directional (one-tailed) tests may be specified should the
empirical situation require it (36, 38).

To demonstrate the utility of \(\hat{Z}_{12}\), we compared \(Z_K\) for
two ecologically-relevant traits in plethodontid salamander (Fig. 4):
surface area to volume ratios (SA:V) and relative body width
(\(\frac{BW}{SVL}\)) (41, 42). While both traits contained significant
phylogenetic signal, tests based on \(\hat{Z}_{12}\) revealed that the
degree of phylogenetic signal was significantly stronger in SA:V
(\(\hat{Z}_{12}=4.13\); \(P=0.000036\): Fig. 4). Biologically, this
observation may be interpreted by the fact that the tropical species --
which form a monophyletic group within plethodontids -- display greater
variation in SA:V, which covaries with disparity in their climatic
niches (42). Thus, greater phylogenetic signal in SA:V is to be
expected.

\begin{figure}
\centering
\includegraphics{fig.4.png}
\caption{(A) Linear measures for relative body size, and regions of the
body used to estimate surface area to volume (SA:V) ratios. (B)
Permutation distributions of phylogenetic signal for SA:V and
\(\frac{BW}{SVL}\), with observed values shown as vertical bars. (C)
Effect sizes (\(Z_K\)) for SA:V and \(\frac{BW}{SVL}\), with their 95\%
confidence intervals (CI not standardized by \(\sqrt(n)\)).{}}
\end{figure}

@@

As an alternative, we described a standardized effect size (\(Z\)) for
assessing the strength of phylogenetic signal. \(Z\) expresses the
magnitude of phylogenetic signal as a standard normal deviate, which is
easily interpretable as the strength of phylogenetic signal relative to
the mean. We applied this concept to both \(\lambda\) and \emph{K}, and
found that \(Z_K\) was a better estimate of the strength of phylogenetic
signal in phenotypic data. First, values of \(Z_K\) more accurately
tracked known changes in the magnitude of phylogenetic signal, as
demonstrated by the near linear relationship between \(Z_K\) and input
signal. Additionally, the precision of \(Z_K\) was more consistent
across the range of input levels of phylogenetic signal. Thus, \(Z_K\)
is a more reliable measure of the relative strength of phylogenetic
signal, and places that effect on a common and comparable scale. We
therefore recommend that future studies interested in evaluating the
strength of phylogenetic signal incorporate \(Z_K\) as a statistical
measure of this effect.

Next we proposed a two-sample test (\(\hat{Z}_{12}\)), which provides a
formal statistical procedure for determining whether the strength of
phylogenetic signal is greater in one phenotypic trait as compared to
another. Prior studies have summarized patterns of variation in
phylogenetic signal across datasets using summary test values, such as
\emph{K} (12). However, because \emph{K} does not scale linearly with
input levels of phylogenetic signal (Fig. 2), and its variance increases
with increasing strength of phylogenetic signal (18, 20), it should not
be considered an effect size that measures the strength of phylogenetic
signal on a common scale. By contrast, standardizing \emph{K} to \(Z_K\)
via equation 2 alleviates these concerns, and facilitates formal
statistical comparisons of the strength of signal across datasets. Thus
when viewed from this perspective, the approach developed here aligns
well with other statistical approaches such as meta-analysis (32, 43,
44), where summary statistics across datasets are converted to
standardized effect sizes for subsequent ``higher order'' statistical
summaries or comparisons. As such, our approach enables evolutionary
biologists to quantitatively examine the relative strength of
phylogenetic signal across a wide range of phenotypic traits, and thus
opens the door for future discoveries that inform on how phenotypic
diversity accumulates in macroevolutionary time across the tree of life.

One important advantage of the approach advocated here is that the
resulting effect sizes (\(Z_K\)) are dimensionless, as the units of
measurement cancel out during the calculation of \(Z\) (45). Thus,
\(Z_K\) represents the strength of phylogenetic signal on a common and
comparable scale -- measured in standard deviations -- regardless of the
initial units and original scale of the phenotypic variables under
investigation. This means that the strength of phylogenetic signal may
be compared across datasets for continuous phenotypic traits measured in
different units and scale, because those units have been standardized
through their conversion to \(Z_K\). For example, our approach could be
utilized to determine whether the strength of phylogenetic signal (say,
in response to ecological differentiation) is stronger in morphological
traits (linear traits: \(mm\)), physiological traits (metabolic rate:
\(\frac{O^2}{min}\)), or behavioral traits (aggression:
\(\frac{\#{displays}}{second}\)). In fact, our empirical example
provided just such a comparison, as SA:V is represented in \(mm^{-1}\)
while relative body size is a unitless ratio (\(\frac{BW}{SVL}\)).
Additionally, our method is capable of comparing the strength of
phylogenetic signal in traits of different dimensionality, as estimates
of phylogenetic signal using \emph{K} have been generalized for
multivariate data (16). Furthermore, tests based on \(\hat{Z}_{12}\) may
be utilized for comparing the strength of phylogenetic signal among
datasets containing a different number of species, and even for
phenotypes obtained from species in different lineages, because their
phylogenetic non-independence and observed variation are taken into
account in the generation of the empirical sampling distribution via
permutation.
